```{r}
# load in libraries used
library(tidyverse)
library(lubridate)
library(readxl)

# read in 3 datasets used
ECdata <- read.csv("EC_ClimateProjections_BC.csv")
BLISS_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
BLISS_unstandardized <- read.csv("Unstandardized_INB.T_OUT.P.csv")
BLISS_unstandardized <- BLISS_unstandardized[complete.cases(BLISS_unstandardized), ]
```

# Forecasting Survivorship - Logic

**Ultimate Goal**: produce a dataset where for each year-scenario pair from 2025 - 2100 we have estimated (i) average minimum temperature of the breeding season, and (ii) precipitation (measured by calculating the total precipitation over a 21-day interval centered around peak egg mass date. From these two estimates, we use the survivorship model to estimate the average adult female survivorship for that year-scenario pair.

1.  **Forecasting the Breeding Season Interval**

-   The breeding season is defined to start on the Bat Lake (BL) ice out date, and end at peak egg mass date.

-   We used GCM mean April temperature to predict ordinal Bat Lake ice out date using simple linear regression ($R^2 = 0.5$).

-   Then, we assumed that peak egg mass date always occurs 22 days after ice out (corresponding to the average observed breeding season length during the 2008-2022 BLISS study period). Using these two estimates, we can easily construct both the breeding interval as well as the precipitation interval.

2.  **Calculating Survivorship within the Intervals**

-   Using these intervals along with daily ENVCAN estimates of minimum temperature and total precipitation, we calculated (i) and (ii) for each year-scenario pair.

## 1. Forecasting the Breeding Interval

```{r}

get_meanAprilTemp_from_EC <- function(yr, scen, dataType="ADJ") {
  # get focal data
  focalData <- ECdata %>% filter(scenario == scen,
                                 year == yr,
                                 month == "Apr")
  # calculate thing
  if (dataType == "ADJ") {
    return(mean(focalData$tasBC))
  }
  if (dataType == "SIM") {
    return(mean(focalData$tas))
  }
  stop("{get_meanAprilTemp_from_EC}: invalid dataType entered.")
}

get_iceOut_from_meanAprilTemp <- function(meanAprilTemp) {
  # SOURCE: Dylan did a regression
  m <- -3.1373
  b <- 125.5190
  return(round((m*meanAprilTemp) + b))
}

mean_breedingSeasonLength <- round(mean(BLISS_study_data$Breeding.Season.Length, na.rm = T))
# assumption: peak egg mass always occurs 22 days after ice out
get_peakEggMass_from_iceOut <- function(iceOut) {
  return(iceOut + mean_breedingSeasonLength)
}
```

## 2. Calculating Survivorship within the Intervals

```{r}

# GET CONSTANTS
obsTmean <- mean(BLISS_unstandardized$T.INB)
obsTsd <- sd(BLISS_unstandardized$T.INB)
obsPmean <- mean(BLISS_unstandardized$P.OUT)
obsPsd <- sd(BLISS_unstandardized$P.OUT)
precipIntervalRadius <- 10 # TODO: why ??????????????????????????????????????
# SOURCE: Dylan's model fitted values
b0 <- 0.9656941 # intercept
b1 <- -0.762042 # standardized temp
b2 <- 0.4460179 # standardized precipitation


# CACLUATE INDICES

get_climateIndices <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  # get filtered data
  tempInterval <- ECdata %>% filter(year == yr,
                                    scenario == scen,
                                    ordinalDay >= iceOut,
                                    ordinalDay <= eggMass)
  precipInterval <- ECdata %>% filter(year == yr,
                                      scenario == scen,
                                      ordinalDay >= eggMass - precipIntervalRadius,
                                      ordinalDay <= eggMass + precipIntervalRadius)
  # in filtered window, calculate indicators: ADJ
  if (dataType == "ADJ") {
    temp = mean(tempInterval$tasminBC)
    wetDays <- precipInterval %>% filter(prBC > 0.0)
    if (nrow(wetDays) == 0) {
      precip <- 0
      warning("{get_climateIndices}: WARNING: a season was found with no wet days.")
    }
    else {
      precip <- sum(precipInterval$prBC) / (nrow(wetDays) * nrow(precipInterval))
    }
    return(c(temp, precip))
  }
  # in filtered window, calculate indicators: SIM
  if (dataType == "SIM") {
    temp <- mean(tempInterval$tasmin)
    wetDays <- precipInterval %>% filter(pr > 0.0)
    if (nrow(wetDays) == 0) {
      precip <- 0
      warning("{get_climateIndices}: WARNING: a season was found with no wet days.")
    }
    else {
      precip = sum(precipInterval$pr) / (nrow(wetDays) * nrow(precipInterval))
    }
    return(c(temp, precip))
  }
  stop("{get_climateIndices}: invalid dataType entered.")
}

get_T <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  val <- get_climateIndices(yr, scen, iceOut, eggMass, dataType)
  return(val[1])
}

get_P <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  val <- get_climateIndices(yr, scen, iceOut, eggMass, dataType)
  return(val[2])
}


# CALCULATE SURVIVORSHIP FROM INDICES

get_survivorship <- function(climateIndices) {
  # standardize temp and precip indices
  standardizedT <- (climateIndices[1] - obsTmean) / obsTsd
  standardizedP <- (climateIndices[2] - obsPmean) / obsPsd
  # estimate mean respionse using logit model
  logitS <- b0 + b1 * standardizedT + b2 * standardizedP
  # convert back to raw probability measure
  S <- 1 / (1 + exp(-logitS))
  # return result
  return(S)
}
```

# Forecasting Survivorship - Impentation

## 1. Creating Final Dataset

```{r}

# For each year, save all of the predicted aspects of the breeding season

final_df <- data.frame(year = integer(),
                       scenario = character(),
                       iceOut = integer(),
                       eggMass = integer(),
                       temperatureIndex = numeric(),
                       precipitationIndex = numeric(),
                       survivorship = numeric(),
                       meanAprilTemp = numeric(), # also save diagnostic information:
                       tempStart = integer(),
                       tempEnd = integer(),
                       precipStart = integer(),
                       precipEnd = integer()
                       )

# setup
years <- 2000:2100
scenarios <- c("SSP126", "SSP245", "SSP585")
data_type <-  "ADJ"
# populate dataframe
for (scen in scenarios) {
  for (yr in years) {
    temp_df <- tibble(year = yr,
                      scenario = scen,
                      tempVar1 = get_meanAprilTemp_from_EC(yr, scen, data_type),
                      iceOut = get_iceOut_from_meanAprilTemp(tempVar1),
                      eggMass = get_peakEggMass_from_iceOut(iceOut),
                      temperatureIndex = get_T(yr, scen, iceOut, eggMass, data_type),
                      precipitationIndex = get_P(yr, scen, iceOut, eggMass, data_type),
                      survivorship = get_survivorship(c(temperatureIndex,precipitationIndex)),
                      meanAprilTemp = tempVar1,
                      tempStart = iceOut,
                      tempEnd = eggMass,
                      precipStart = eggMass - precipIntervalRadius,
                      precipEnd = eggMass + precipIntervalRadius
                      )
    temp_df <- temp_df %>% select(-tempVar1)
    final_df <- rbind(final_df, temp_df)
  }
}
```

```{r}
# uncomment to write to CSV
# write_csv(final_df, "forecasted_survivorship.csv")
```

## 2. Diagnostic Plots

```{r}

final_df <- read.csv("forecasted_survivorship.csv")

plottingData5 <- final_df %>% filter(scenario == "SSP585")
plottingData1 <- final_df %>% filter(scenario == "SSP126")
plottingData2 <- final_df %>% filter(scenario == "SSP245")


# BAT LAKE ICE OUT DATE

plot(x = 2000:2100, y = plottingData5$iceOut, 
     ylim = c(90, 130), type = "l", col = "darkred", ylab = "Ice Out Date", xlab = "")
lines(x = 2008:2022, y = BLISS_study_data$Ice.Out.Ordinal)
lines(x = 2008:2022, y = rep(mean(BLISS_study_data$Ice.Out.Ordinal), 15), col = "grey")
lines(x = 2000:2100, y = plottingData1$iceOut, col = adjustcolor("forestgreen", alpha.f = 0.4))
lines(x = 2000:2100, y = plottingData2$iceOut, col = adjustcolor("navyblue", alpha.f = 0.4))

# TEMPERATURE INDEX

plot(x = 2000:2100, y = plottingData5$temperatureIndex,
     ylim = c(-2, 4), type = "l", col = "darkred", ylab = "Temperature Index", xlab = "")
lines(x = 2008:2022, y = BLISS_unstandardized$T.INB)
lines(x = 2008:2022, y = rep(mean(BLISS_unstandardized$T.INB), 15), col = "grey")
lines(x = 2000:2100, y = plottingData1$temperatureIndex,
      col = adjustcolor("forestgreen", alpha.f = 0.4))
lines(x = 2000:2100, y = plottingData2$temperatureIndex,
      col = adjustcolor("navyblue", alpha.f = 0.4))

# PRECIPITATION INDEX

plot(x = 2000:2100, y = plottingData5$precipitationIndex,
     ylim = c(0, 0.6), type = "l", col = "darkred", ylab = "Precipitation Index", xlab = "")
lines(x = 2008:2022, y = BLISS_unstandardized$P.OUT, lw = 2)
lines(x = 2008:2022, y = rep(mean(BLISS_unstandardized$P.OUT), 15), col = "grey")
lines(x = 2000:2100, y = plottingData1$precipitationIndex,
      col = adjustcolor("forestgreen", alpha.f = 0.4))
lines(x = 2000:2100, y = plottingData2$precipitationIndex,
      col = adjustcolor("navyblue", alpha.f = 0.4))

# SURVIVORSHIP

BLISS_survivorship <- as.numeric(apply(BLISS_unstandardized[, c(2, 3)], 1, get_survivorship))

plot(x = 2000:2100, y = plottingData5$survivorship,
     ylim = c(0.25, 0.9), type = "l", col = "darkred", ylab = "Annual Survivorship", xlab = "")
# lines(x = 2008:2022, y = BLISS_survivorship, lw = 2)
lines(x = 2008:2022, y = rep(mean(BLISS_survivorship), 15), col = "grey")
lines(x = 2000:2100, y = rep(0.7242604, 101), col = "red", lty = "dotted")
lines(x = 2000:2100, y = plottingData1$survivorship,
      col = adjustcolor("forestgreen", alpha.f = 0.4))
lines(x = 2000:2100, y = plottingData2$survivorship,
      col = adjustcolor("navyblue", alpha.f = 0.4))
```

# Forecasting Population

## 1. Population Growth Model

```{r}
library(popbio)

# DEFINING PARAMETERS

age_at_maturity <- 6 # SRC: Boyle - skeletochronology
adult_fem_abundance <- 1840 # SRC: Moldowan, estimated in 2019
adult_fem_abundance2 <- 1831 # SRC: Moldowan, estimated directly from figure
s_a <- 1 / (1 + exp(-0.9656941)) # SRC: Dylan's intercept
f <- 1.41 # we know Patrick's estimates are underestimates, so works nicely with an overestimated p_skip
p_skip <- 0.5
s_m <- 0.8551498 # optimized from other parameters such that lambda = 1

get_rescaled_vector <- function(some_vector, k) {
  n_stable_fem <- some_vector[length(some_vector)]
  factor <- k / n_stable_fem
  return(factor * some_vector)
}

# CONSTRUCTING PROJECTION MATRIX

dimension <- age_at_maturity + 1 # ages 0-1, ..., 6-7, and 7+
model1_matrix <- matrix(0, dimension, dimension)
model1_matrix[1, dimension] <- f*(1-p_skip)
model1_matrix[dimension, dimension] <- s_a
for (i in 1:(dimension - 1)) {
  model1_matrix[i+1, i] <- s_m
}

eigen_info <- eigen.analysis(model1_matrix)
lambda1 <- eigen_info$lambda1
r1 <- log(lambda1)
stable_distribution <- eigen_info$stable.stage
stable_distribution_cal <- get_rescaled_vector(stable_distribution,
                                               adult_fem_abundance)

# SIMULATION HELPER FUNCTIONS

get_variant <- function(s) {
  result <- model1_matrix
  result[dimension, dimension] <- s
  return(result)
}
```

## 2. Simulation

```{r}

# setup
final_df <- read.csv("forecasted_survivorship.csv")
set.seed(416)
years_considered <- 2019:2100
envData <- final_df %>% filter(year %in% years_considered)

# SIMULATION FUNCTION

data_used <- envData
project_population <- function(scen) {
  # filter current scenario
  filteredData <- data_used %>% filter(scenario == scen) %>% arrange(year)
  # initialize pop vector
  init_N <- stable_distribution_cal
  # simulation setup
  curr_N <- init_N
  survivorships <- filteredData$survivorship
  pop_data <- numeric(length(years_considered))
  pop_data[1] <- curr_N[dimension]
  #simulation
  for (i in 1:(length(years_considered)-1)) {
    # update matrix with adult female survival
    curr_model_variant <- get_variant(survivorships[i])
    # do the simulatuion as normal
    curr_N <- curr_model_variant %*% curr_N
    # save the number of adult females in the current population vector
    pop_data[i+1] <- curr_N[dimension]
  }
  return(pop_data)
}

# PLOTTING

plotting_data <- data.frame(x_axis = years_considered,
                            y5 = project_population("SSP585"),
                            y1 = project_population("SSP126"),
                            y2 = project_population("SSP245"))

plotting_data %>%
  pivot_longer(cols = starts_with("y"), names_to = "scenario", values_to = "abundance") %>%
  ggplot(aes(x = x_axis, y = abundance, color = scenario)) +
  geom_line(alpha = 0.75) +
#  geom_hline(yintercept = adult_fem_abundance, linetype = "dashed", color = "grey30") +
#    geom_hline(yintercept = adult_fem_abundance / 2, linetype = "dashed", color = "grey50") +
#  geom_vline(xintercept = 2026, linetype = "dashed", color = "grey30") +
  scale_color_manual(values = c("y1" = "forestgreen", "y2" = "navyblue", "y5" = "darkred"),
                     labels = c("y1" = "SSP1-2.6", "y2" = "SSP2-4.5", "y5" = "SSP5-8.5")) +
  coord_cartesian(ylim = c(0, 2500)) +
  theme_classic() +
  theme(
    legend.position = "top", legend.title = element_blank(),
    axis.title = element_text(face = "bold", size = 10)) +
  labs(x = "Time (Years)", y = "Adult Female Abundance")

```
