## Setup

```{r}
library(tidyverse)
library(lubridate)
library(itsmr)
library(broom)
```

## Data Cleaning

```{r}

# read in raw data
rawAlg <- read.csv("Algonquin East Gate Climate Data.csv")
rawEC <- read.csv("ECCC_ClimateProjections.csv")

# preliminary cleaning
cleanAlg <- rawAlg %>% 
  select(
  LOCAL_DATE, LOCAL_DAY, LOCAL_YEAR, LOCAL_MONTH,
  MEAN_TEMPERATURE, MIN_TEMPERATURE, TOTAL_PRECIPITATION
  ) %>%
  mutate(decimalDate = decimal_date(ymd(LOCAL_DATE))) %>%
  filter(LOCAL_MONTH %in% c(3, 4, 5)) # this also removes NA rows

cleanEC <- rawEC %>% 
  filter(year %in% 2005:2024) %>%
  filter(scenario == "SSP585") %>% # only using SSP585 to determie correction factors
  mutate(decimalDate = decimal_date(ymd(date))) %>%
  arrange(decimalDate)

# check that all rows are unique dates
nrow(cleanEC) == length(unique(cleanEC$date))
nrow(cleanAlg) == length(unique(cleanAlg$LOCAL_DATE))
# find the dates that are missing from Alg data from 2005-2024
missing_dates <- lubridate::setdiff(unique(cleanEC$date), unique(cleanAlg$LOCAL_DATE))
# remove them from EC
cleanEC <- cleanEC %>% filter(!(date %in% missing_dates))
# final checks of date accordance
all(cleanEC$date == cleanAlg$LOCAL_DATE)
all(cleanEC$decimalDate == cleanAlg$decimalDate)

# plot the data for a first peek
yearPlot <- 2008


plot2005EC <- cleanEC %>% filter(year== yearPlot)
plot2005Alg <- cleanAlg %>% filter(LOCAL_YEAR == yearPlot)

plot(x = plot2005EC$decimalDate, y = plot2005EC$tas,
     type = "l", col = "cornflowerblue", ylim = c(-20, 20))
lines(x = plot2005EC$decimalDate, y = plot2005Alg$MEAN_TEMPERATURE)
lines(x = plot2005EC$decimalDate, y = smooth.ma(plot2005Alg$MEAN_TEMPERATURE, 10),
      col = "grey")

```

## ISI-MIP approach

1.  Bias Correction

    1.  Add constant offset C to each point for tas and tasmin, representing systematic, long-term bias in the simulated temperature data
    2.  Add constant multiplicative offset c to each point in pr, prepresenting systematic, long-term bias in the simulated precipitation data

    Note: C is different for each month

2.  Variance Inflation

    1.  

Note that our corrections are applied to SSP585 (status-quo model) and assumed to be similar for the other datasets (although the validity of this assumption can easily be verified by changing one line in the existing code).

## 1. Bias Correction

```{r}

monthEC <- cleanEC %>% group_by(year, month) %>% summarise(
  days = n(),
  meanPr = mean(pr),
  meanTas = mean(tas),
  meanTasmin = mean(tasmin)
  ) %>% arrange(year, month)
#seasonEC <- cleanEC %>% group_by(year) %>% summarise(
#  days = n(),
#  meanPr = mean(pr),
#  meanTas = mean(tas),
#  meanTasmin = mean(tasmin)
#)

monthAlg <- cleanAlg %>% group_by(LOCAL_YEAR, LOCAL_MONTH) %>% summarise(
  days = n(),
  propNAPr = sum(is.na(TOTAL_PRECIPITATION)) / days,
  propNATas = sum(is.na(MEAN_TEMPERATURE)) / days,
  propNATasmin = sum(is.na(MIN_TEMPERATURE)) / days,
  usePr = ifelse(propNAPr > 0.2, FALSE, TRUE),
  useTas = ifelse(propNATas > 0.2, FALSE, TRUE),
  useTasmin = ifelse(propNATasmin > 0.2, FALSE, TRUE),
  meanPr = mean(TOTAL_PRECIPITATION, na.rm = TRUE),
  meanTas = mean(MEAN_TEMPERATURE, na.rm = TRUE),
  meanTasmin = mean(MIN_TEMPERATURE, na.rm = TRUE)
  ) %>% arrange(LOCAL_YEAR, LOCAL_MONTH) %>% 
  mutate(LOCAL_MONTH = case_when(
    LOCAL_MONTH == 3 ~ "Mar",
    LOCAL_MONTH == 4 ~ "Apr",
    LOCAL_MONTH == 5 ~"May")
    )
#seasonAlg <- cleanAlg %>% group_by(LOCAL_YEAR) %>% summarise(
#  days = n(),
#  propNAPr = sum(is.na(TOTAL_PRECIPITATION)) / days,
#  propNATas = sum(is.na(MEAN_TEMPERATURE)) / days,
#  propNATasmin = sum(is.na(MIN_TEMPERATURE)) / days,
#  usePr = ifelse(propNAPr > 0.2, FALSE, TRUE),
#  useTas = ifelse(propNATas > 0.2, FALSE, TRUE),
#  useTasmin = ifelse(propNATasmin > 0.2, FALSE, TRUE),
#  meanPr = mean(TOTAL_PRECIPITATION, na.rm = TRUE),
#  meanTas = mean(MEAN_TEMPERATURE, na.rm = TRUE),
#  meanTasmin = mean(MIN_TEMPERATURE, na.rm = TRUE)
#)

# graph use... onto EC data:
monthEC <- monthEC %>%
  left_join(monthAlg %>% 
              select(LOCAL_YEAR, LOCAL_MONTH, useTas, useTasmin, usePr),
            by = c("year" = "LOCAL_YEAR", "month" = "LOCAL_MONTH"))

# calculate temperature offsets
calculate_Tas_offsets <- function() {
  # return vector
  returnVector <- numeric(3)
  # filter datasets
  EC3 <- monthEC %>% filter(month == "Mar") %>% filter(useTas)
  Alg3 <- monthAlg %>% filter(LOCAL_MONTH == "Mar") %>% filter(useTas)
  EC4 <- monthEC %>% filter(month == "Apr")  %>% filter(useTas)
  Alg4 <- monthAlg %>% filter(LOCAL_MONTH == "Apr") %>% filter(useTas)
  EC5 <- monthEC %>% filter(month == "May")  %>% filter(useTas)
  Alg5 <- monthAlg %>% filter(LOCAL_MONTH == "May") %>% filter(useTas)
  # cacluate Tas offsets
  returnVector[1] = mean(Alg3$meanTas - EC3$meanTas)
  returnVector[2] = mean(Alg4$meanTas - EC4$meanTas)
  returnVector[3] = mean(Alg5$meanTas - EC5$meanTas)
  # return
  return(returnVector)
}
calculate_Tasmin_offsets <- function() {
  # return vector
  returnVector <- numeric(3)
  # filter datasets
  EC3 <- monthEC %>% filter(month == "Mar")  %>% filter(useTasmin)
  Alg3 <- monthAlg %>% filter(LOCAL_MONTH == "Mar") %>% filter(useTasmin)
  EC4 <- monthEC %>% filter(month == "Apr")  %>% filter(useTasmin)
  Alg4 <- monthAlg %>% filter(LOCAL_MONTH == "Apr") %>% filter(useTasmin)
  EC5 <- monthEC %>% filter(month == "May")  %>% filter(useTasmin)
  Alg5 <- monthAlg %>% filter(LOCAL_MONTH == "May") %>% filter(useTasmin)
  # cacluate Tas offsets
  returnVector[1] = mean(Alg3$meanTasmin - EC3$meanTasmin)
  returnVector[2] = mean(Alg4$meanTasmin - EC4$meanTasmin)
  returnVector[3] = mean(Alg5$meanTasmin - EC5$meanTasmin)
  # return
  return(returnVector)
}

# calculate precipitation offsets
calculate_Pr_offsets <- function() {
  # return vector
  returnVector <- numeric(3)
  # filter datasets
  EC3 <- monthEC %>% filter(month == "Mar")  %>% filter(usePr)
  Alg3 <- monthAlg %>% filter(LOCAL_MONTH == "Mar") %>% filter(usePr)
  EC4 <- monthEC %>% filter(month == "Apr")  %>% filter(usePr)
  Alg4 <- monthAlg %>% filter(LOCAL_MONTH == "Apr") %>% filter(usePr)
  EC5 <- monthEC %>% filter(month == "May")  %>% filter(useTasmin)
  Alg5 <- monthAlg %>% filter(LOCAL_MONTH == "May") %>% filter(usePr)
  # cacluate precip offsets
  returnVector[1] <- sum(Alg3$meanPr) / sum(EC3$meanPr)
  returnVector[2] <- sum(Alg4$meanPr) / sum(EC4$meanPr)
  returnVector[3] <- sum(Alg5$meanPr) / sum(EC5$meanPr)
  # return
  return(returnVector)
}

monthlyTasOffsets <-  calculate_Tas_offsets()
monthlyTasminOffsets <- calculate_Tasmin_offsets()
monthlyPrOffsets <- calculate_Pr_offsets()
```

## Apply Bias and Validate with Plots

```{r}
# apply bias corrections to full ECCC dataset
biasCorrectedEC <- rawEC %>% 
  mutate(pr = case_when(month == "Mar" ~ monthlyPrOffsets[1] * pr,
                        month == "Apr" ~ monthlyPrOffsets[2] * pr,
                        month == "May" ~ monthlyPrOffsets[3] * pr)
         ) %>%
  mutate(tas = case_when(month == "Mar" ~ monthlyTasOffsets[1] + tas,
                         month == "Apr" ~ monthlyTasOffsets[2] + tas,
                         month == "May" ~ monthlyTasOffsets[3] + tas)
         ) %>%
  mutate(tasmin = case_when(month == "Mar" ~ monthlyTasminOffsets[1] + tasmin,
                            month == "Apr" ~ monthlyTasminOffsets[2] + tasmin,
                            month == "May" ~ monthlyTasminOffsets[3] + tasmin))

# check that the data looks to be bias-corrected
cleanedBiasedEC <- biasCorrectedEC %>% 
  filter(year %in% 2005:2024) %>%
  filter(scenario == "SSP585") %>%
  mutate(decimalDate = decimal_date(ymd(date))) %>%
  arrange(decimalDate)
missing_dates2 <- setdiff(unique(cleanedBiasedEC$date), unique(cleanAlg$LOCAL_DATE))
cleanedBiasedEC <- cleanedBiasedEC %>% filter(!(date %in% missing_dates2))

yearPlot <- 2014
plot2005debiasedEC <- cleanedBiasedEC %>% filter(year== yearPlot)
plot2005EC <- cleanEC %>% filter(year== yearPlot)
plot2005Alg <- cleanAlg %>% filter(LOCAL_YEAR == yearPlot)

# PLOT MEAN TEMP
plot(x = plot2005EC$decimalDate, y = plot2005debiasedEC$tas,
     type = "l", col = "navyblue", ylim = c(-30, 25),
     xlab = "Time (Year)", ylab = "Mean Daily Temperature")
lines(x = plot2005EC$decimalDate, y = plot2005EC$tas, col = "cornflowerblue")
lines(x = plot2005EC$decimalDate, y = plot2005Alg$MEAN_TEMPERATURE)
lines(x = plot2005EC$decimalDate, y = smooth.ma(plot2005Alg$MEAN_TEMPERATURE, 10),
      col = "grey")
# PLOT MIN TEMP
plot(x = plot2005EC$decimalDate, y = plot2005debiasedEC$tasmin,
     type = "l", col = "navyblue", ylim = c(-30, 25),
     xlab = "Time (Year)", ylab = "Min Daily Temperature")
lines(x = plot2005EC$decimalDate, y = plot2005EC$tasmin, col = "cornflowerblue")
lines(x = plot2005EC$decimalDate, y = plot2005Alg$MIN_TEMPERATURE)
lines(x = plot2005EC$decimalDate, y = smooth.ma(plot2005Alg$MIN_TEMPERATURE, 10),
      col = "grey")
# PLOT PRECIP
plot(x = plot2005EC$decimalDate, y = plot2005debiasedEC$pr,
     type = "l", col = "navyblue", ylim = c(-30, 25),
     xlab = "Time (Year)", ylab = "Daily Precipitation (Total Amount)")
lines(x = plot2005EC$decimalDate, y = plot2005EC$pr, col = "cornflowerblue")
lines(x = plot2005EC$decimalDate, y = plot2005Alg$TOTAL_PRECIPITATION)
lines(x = plot2005EC$decimalDate, y = smooth.ma(plot2005Alg$TOTAL_PRECIPITATION, 10),
      col = "grey")

# save resulting data to a new .csv file
# write_csv(biasCorrectedEC, "ECCC_ClimateProjections_BiasCorrected.csv")
```

## 2. Variance Inflation

### 1. Preparing Data

```{r}
# import the bias-corrected data
biasCorrectedEC <- read.csv("ECCC_ClimateProjections_BiasCorrected.csv")
rawAlg <- read.csv("Algonquin East Gate Climate Data.csv")

# EXACT SAME PRELIMINARY CLEANING PROCEDURE AS ABOVE

cleanAlg <- rawAlg %>% 
  select(
  LOCAL_DATE, LOCAL_DAY, LOCAL_YEAR, LOCAL_MONTH,
  MEAN_TEMPERATURE, MIN_TEMPERATURE, TOTAL_PRECIPITATION
  ) %>%
  mutate(decimalDate = decimal_date(ymd(LOCAL_DATE))) %>%
  filter(LOCAL_MONTH %in% c(3, 4, 5)) %>% # this also removes NA rows
  mutate(LOCAL_MONTH = case_when(
    LOCAL_MONTH == 3 ~ "Mar",
    LOCAL_MONTH == 4 ~ "Apr",
    LOCAL_MONTH == 5 ~"May")
    )

cleanEC <- biasCorrectedEC %>% 
  filter(year %in% 2005:2024) %>%
  filter(scenario == "SSP585") %>% # only using SSP585 to determie correction factors
  mutate(decimalDate = decimal_date(ymd(date))) %>%
  arrange(decimalDate)

# check that all rows are unique dates
nrow(cleanEC) == length(unique(cleanEC$date))
nrow(cleanAlg) == length(unique(cleanAlg$LOCAL_DATE))
# find the dates that are missing from Alg data from 2005-2024
missing_dates3 <- lubridate::setdiff(unique(cleanEC$date), unique(cleanAlg$LOCAL_DATE))
# remove them from EC
cleanEC <- cleanEC %>% filter(!(date %in% missing_dates3))
# final checks of date accordance
all(cleanEC$date == cleanAlg$LOCAL_DATE)
all(cleanEC$decimalDate == cleanAlg$decimalDate)

# ADD MONTHLY RESIDUALS (NORMALIZED FORMS for precipitation) TO CLEAN EC DATA

monthEC <- cleanEC %>% group_by(year, month) %>% summarise(
  days = n(),
  meanPr = mean(pr),
  meanTas = mean(tas),
  meanTasmin = mean(tasmin)
  ) %>% arrange(year, month)

cleanEC <- cleanEC %>% 
  left_join(monthEC %>% select(year, month, meanTas, meanTasmin, meanPr),
            by = c("year", "month")) %>%
  mutate(residTas = tas - meanTas) %>%
  mutate(residTasmin = tasmin - meanTasmin) %>% 
  mutate(normPr = pr / meanPr) # use a different formula because of positivity constraint

# ADD MONTHLY RESIDUALS (NORMALIZED FORMS for precipitation) TO CLEAN ALG DATA
# Note: currently ignoring NAs completely

monthAlg <- cleanAlg %>% group_by(LOCAL_YEAR, LOCAL_MONTH) %>% summarise(
  days = n(),
  propNAPr = sum(is.na(TOTAL_PRECIPITATION)) / days,
  propNATas = sum(is.na(MEAN_TEMPERATURE)) / days,
  propNATasmin = sum(is.na(MIN_TEMPERATURE)) / days,
  usePr = ifelse(propNAPr > 0.2, FALSE, TRUE),
  useTas = ifelse(propNATas > 0.2, FALSE, TRUE),
  useTasmin = ifelse(propNATasmin > 0.2, FALSE, TRUE),
  meanPr = mean(TOTAL_PRECIPITATION, na.rm = TRUE),
  meanTas = mean(MEAN_TEMPERATURE, na.rm = TRUE),
  meanTasmin = mean(MIN_TEMPERATURE, na.rm = TRUE)
  ) %>% arrange(LOCAL_YEAR, LOCAL_MONTH)

cleanAlg <- cleanAlg %>% 
  left_join(monthAlg %>% select(LOCAL_YEAR, LOCAL_MONTH, meanTas, meanTasmin, meanPr),
            by = c("LOCAL_YEAR", "LOCAL_MONTH")) %>%
  mutate(residTas = MEAN_TEMPERATURE - meanTas) %>%
  mutate(residTasmin = MIN_TEMPERATURE - meanTasmin) %>% 
  mutate(normPr = TOTAL_PRECIPITATION / meanPr)
```

### 2. Temperature Inflation

```{r}

# plot temp resids
x <- cleanAlg %>% filter(LOCAL_MONTH == "May")
xx <- cleanEC %>% filter(month == "May")
plot(xx$residTas, x$residTas)
abline(lm(x$residTas ~ xx$residTas - 1))
hist(xx$residTas)
hist(x$residTas)


# APPLY PARAMETRIC QUANTILE MAPPING TO TEMPERATURE RESIDUALS
# Note: if monthly varability is independent of monthly mean, then the fitted transfer function should be applicable to any data, regardless of it's temporal range
# Note: PARAMETRIC quantile mapping is required so that the fitted transfer function can be applied to future EC data

# now run 3 linear regressions to determine the month-specific slopes of the transfer function
slopes_table <- cleanAlg %>%
  inner_join(cleanEC, by = "decimalDate") %>%
  filter(month %in% c("Mar", "Apr", "May")) %>%
  group_by(month) %>%
  summarise(
    slopeTas = coef(lm(residTas.x ~ residTas.y)),
    slopeTasmin = coef(lm(residTasmin.x ~ residTasmin.y))
  )
# Note: R^2 for the models are around 20, and p-values are always small

```
