```{r}
# load in libraries used
library(tidyverse)
library(lubridate)
library(readxl)

# read in 3 datasets used
ECdata <- read.csv("EC_ClimateProjections_BC.csv")
BLISS_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
BLISS_unstandardized <- read.csv("Unstandardized_INB.T_OUT.P.csv")
BLISS_unstandardized <- BLISS_unstandardized[complete.cases(BLISS_unstandardized), ]
```

# Forecasting Survivorship

**Ultimate Goal**: produce a dataset where for each year-scenario pair from 2025 - 2100 we have estimated (i) average minimum temperature of the breeding season, and (ii) precipitation (measured by calculating the total precipitation over a 21-day interval centered around peak egg mass date. From these two estimates, we use the survivorship model to estimate the average adult female survivorship for that year-scenario pair.

**Some Information:**

-   The breeding season is defined to start on the Bat Lake (BL) ice out date, and end at peak egg mass date.

-   We used mean April temperature to predict Bat Lake ice out date using simple linear regression.

-   Then, we assumed that peak egg mass date always occurs 22 days after ice out (corresponding to the average observed breeding season length during the 2008-2022 BLISS study period). Using these two estimates, we can easily construct both the breeding interval as well as the precipitation interval.

-   Using these intervals along with daily ENVCAN estimates of minimum temperature and total precipitation, we calculated (i) and (ii) for each year-scenario pair.

-   *Note: all dates are measured and used ordinally*

**Regression information:**

-   BL ice out date is predicted well ($R^2 = 0.5$) by mean April temperature (Note: estimates taken from a screenshot of the summary output of a fitted simple regression model).

## 1. Forecasting Intervals

### 1.1. Predicting Bat Lake Ice Out From GCM Mean April Temperature

```{r}

get_meanAprilTemp_from_EC <- function(yr, scen, dataType="ADJ") {
  # get focal data
  focalData <- ECdata %>% filter(scenario == scen,
                                 year == yr,
                                 month == "Apr")
  # calculate thing
  if (dataType == "ADJ") {
    return(mean(focalData$tasBC))
  }
  if (dataType == "SIM") {
    return(mean(focalData$tas))
  }
  stop("{get_meanAprilTemp_from_EC}: invalid dataType entered.")
}

get_iceOut_from_meanAprilTemp <- function(meanAprilTemp) {
  # SOURCE: Dylan did a regression (iMessage screenshot)
  m <- -3.1373
  b <- 125.5190
  return(round((m*meanAprilTemp) + b))
}

# get ice out estimates
iceOutEstimatesADJ <- numeric()
iceOutEstimatesSIM <- numeric()
for (yr in 2000:2100) {
  currMATADJ <- get_meanAprilTemp_from_EC(yr, "SSP585", "ADJ")
  currMATSIM <- get_meanAprilTemp_from_EC(yr, "SSP585", "SIM")
  curriceOutADJ <- get_iceOut_from_meanAprilTemp(currMATADJ)
  curriceOutSIM <- get_iceOut_from_meanAprilTemp(currMATSIM)
  iceOutEstimatesADJ <- c(iceOutEstimatesADJ, curriceOutADJ)
  iceOutEstimatesSIM <- c(iceOutEstimatesSIM, curriceOutSIM)
}

# compare with true observations
realiceOut <- numeric(101)
realiceOut[1:8] <- NA
realiceOut[9:23] <- BLISS_study_data$Ice.Out.Ordinal
realiceOut[24:101] <- NA

plot(x = 2000:2040, y = iceOutEstimatesADJ[1:41],
     type = "l", ylim = c(100, 135), col = "navy", xlab = "", ylab = "Ice Out (Ordinal)")
lines(x = 2000:2040, y = iceOutEstimatesSIM[1:41], col = "cornflowerblue")
lines(x = 2000:2040, y = realiceOut[1:41], lw = 2)
lines(x = 2000:2040, y = rep(mean(realiceOut, na.rm = T), 41), col = "grey")
abline(v = c(2008, 2022), col = "darkred", lty = 2, lwd = 1.5)

```

Plot shows that applying bias correction to mean temperature also corrected the systematically early ice outs predicted by the unaltered GCM.

### 1.2 Predicting Peak Egg Mass from Bat Lake Ice Out

```{r}

# setup
mean_breedingSeasonLength <- round(mean(BLISS_study_data$Breeding.Season.Length, na.rm = T))
# assumption: peak egg mass always occurs 22 days after ice out
get_peakEggMass_from_iceOut <- function(iceOut) {
  return(iceOut + mean_breedingSeasonLength)
}

# get peak egg mass estimates
peakEggMassEstimatesADJ <- get_peakEggMass_from_iceOut(iceOutEstimatesADJ) 
peakEggMassEstimatesSIM <- get_peakEggMass_from_iceOut(iceOutEstimatesSIM) 

# compare with true observations
#realpeakEggMass <- numeric(101)
#realpeakEggMass[1:8] <- NA
#realpeakEggMass[9:23] <- BLISS_study_data$Peak.Egg.Mass.Ordinal
#realpeakEggMass[24:101] <- NA

#plot(x = 2000:2040, y = peakEggMassEstimatesADJ[1:41],
#     type = "l", ylim = c(120, 150), col = "navy", xlab = "", ylab = "Peak Egg Mass (Ordinal)")
#lines(x = 2000:2040, y = peakEggMassEstimatesSIM[1:41], col = "cornflowerblue")
#lines(x = 2000:2040, y = realpeakEggMass[1:41], lw = 2)
#lines(x = 2000:2040, y = rep(mean(realpeakEggMass, na.rm = T), 41), col = "grey")
#abline(v = c(2008, 2022), col = "darkred", lty = 2, lwd = 1.5)
```

## 2. Calculations within Intervals

### 2.1 Logic Functions

```{r}

# GET CONSTANTS
obsTmean <- mean(BLISS_unstandardized$T.INB)
obsTsd <- sd(BLISS_unstandardized$T.INB)
obsPmean <- mean(BLISS_unstandardized$P.OUT)
obsPsd <- sd(BLISS_unstandardized$P.OUT)
precipIntervalRadius <- 10 # TODO: why ??????????????????????????????????????
# SOURCE: Dylan's model fitted values
b0 <- 0.9656941 # intercept
b1 <- -0.762042 # standardized temp
b2 <- 0.4460179 # standardized precipitation


# CACLUATE INDICES

get_climateIndices <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  # get filtered data
  tempInterval <- ECdata %>% filter(year == yr,
                                    scenario == scen,
                                    ordinalDay >= iceOut,
                                    ordinalDay <= eggMass)
  precipInterval <- ECdata %>% filter(year == yr,
                                      scenario == scen,
                                      ordinalDay >= eggMass - precipIntervalRadius,
                                      ordinalDay <= eggMass + precipIntervalRadius)
  # in filtered window, calculate indicators: ADJ
  if (dataType == "ADJ") {
    temp = mean(tempInterval$tasminBC)
    wetDays <- precipInterval %>% filter(prBC > 0.0)
    if (nrow(wetDays) == 0) {
      precip <- 0
      warning("{get_climateIndices}: WARNING: a season was found with no wet days.")
    }
    else {
      precip <- sum(precipInterval$prBC) / (nrow(wetDays) * nrow(precipInterval))
    }
    return(c(temp, precip))
  }
  # in filtered window, calculate indicators: SIM
  if (dataType == "SIM") {
    temp <- mean(tempInterval$tasmin)
    wetDays <- precipInterval %>% filter(pr > 0.0)
    if (nrow(wetDays) == 0) {
      precip <- 0
      warning("{get_climateIndices}: WARNING: a season was found with no wet days.")
    }
    else {
      precip = sum(precipInterval$pr) / (nrow(wetDays) * nrow(precipInterval))
    }
    return(c(temp, precip))
  }
  stop("{get_climateIndices}: invalid dataType entered.")
}

get_T <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  val <- get_climateIndices(yr, scen, iceOut, eggMass, dataType)
  return(val[1])
}

get_P <- function(yr, scen, iceOut, eggMass, dataType="ADJ") {
  val <- get_climateIndices(yr, scen, iceOut, eggMass, dataType)
  return(val[2])
}


# CALCULATE SURVIVORSHIP FROM INDICES

get_survivorship <- function(climateIndices) {
  # standardize temp and precip indices
  standardizedT <- (climateIndices[1] - obsTmean) / obsTsd
  standardizedP <- (climateIndices[2] - obsPmean) / obsPsd
  # estimate mean respionse using logit model
  logitS <- b0 + b1 * standardizedT + b2 * standardizedP
  # convert back to raw probability measure
  S <- 1 / (1 + exp(-logitS))
  # return result
  return(S)
}
```

### 2.2 Create Final Dataset

```{r}

# For each year, save all of the predicted aspects of the breeding season

final_df <- data.frame(year = integer(),
                       scenario = character(),
                       iceOut = integer(),
                       eggMass = integer(),
                       temperatureIndex = numeric(),
                       precipitationIndex = numeric(),
                       survivorship = numeric(),
                       meanAprilTemp = numeric(), # also save diagnostic information:
                       tempStart = integer(),
                       tempEnd = integer(),
                       precipStart = integer(),
                       precipEnd = integer()
                       )

# setup
years <- 2000:2100
scenarios <- c("SSP585")
data_type <-  "ADJ"
# populate dataframe
for (scen in scenarios) {
  for (yr in years) {
    temp_df <- tibble(year = yr,
                      scenario = scen,
                      tempVar1 = get_meanAprilTemp_from_EC(yr, scen, data_type),
                      iceOut = get_iceOut_from_meanAprilTemp(tempVar1),
                      eggMass = get_peakEggMass_from_iceOut(iceOut),
                      temperatureIndex = get_T(yr, scen, iceOut, eggMass, data_type),
                      precipitationIndex = get_P(yr, scen, iceOut, eggMass, data_type),
                      survivorship = get_survivorship(c(temperatureIndex,precipitationIndex)),
                      meanAprilTemp = tempVar1,
                      tempStart = iceOut,
                      tempEnd = eggMass,
                      precipStart = eggMass - precipIntervalRadius,
                      precipEnd = eggMass + precipIntervalRadius
                      )
    temp_df <- temp_df %>% select(-tempVar1)
    final_df <- rbind(final_df, temp_df)
  }
}

# datset output tests
all(final_df$iceOut == iceOutEstimatesADJ)
all(final_df$eggMass == peakEggMassEstimatesADJ)

# BAT LAKE ICE OUT DATE

plot(x = 2000:2100, y = final_df$iceOut, 
     ylim = c(90, 130), type = "l", col = "navy", ylab = "Ice Out Date", xlab = "")
lines(x = 2008:2022, y = BLISS_study_data$Ice.Out.Ordinal)
lines(x = 2008:2022, y = rep(mean(BLISS_study_data$Ice.Out.Ordinal), 15), col = "grey")

# TEMPERATURE INDEX

plot(x = 2000:2100, y = final_df$temperatureIndex,
     ylim = c(-2, 4), type = "l", col = "navy", ylab = "Temperature Index", xlab = "")
lines(x = 2008:2022, y = BLISS_unstandardized$T.INB)
lines(x = 2008:2022, y = rep(mean(BLISS_unstandardized$T.INB), 15), col = "grey")

# PRECIPITATION INDEX

plot(x = 2000:2100, y = final_df$precipitationIndex,
     ylim = c(0, 0.6), type = "l", col = "navy", ylab = "Precipitation Index", xlab = "")
lines(x = 2008:2022, y = BLISS_unstandardized$P.OUT, lw = 2)
lines(x = 2008:2022, y = rep(mean(BLISS_unstandardized$P.OUT), 15), col = "grey")

# SURVIVORSHIP

BLISS_survivorship <- as.numeric(apply(BLISS_unstandardized[, c(2, 3)], 1, get_survivorship))

plot(x = 2000:2100, y = final_df$survivorship,
     ylim = c(0.25, 0.9), type = "l", col = "navy", ylab = "Annual Survivorship", xlab = "")
# lines(x = 2008:2022, y = BLISS_survivorship, lw = 2)
lines(x = 2008:2022, y = rep(mean(BLISS_survivorship), 15), col = "grey")
lines(x = 2000:2100, y = rep(0.7242604, 101), col = "red")

# write to CSV
# write_csv(final_df, "final_data.csv")
```

# Forecasting Population

## 1. Population Growth Model

```{r}
library(popbio)

# DEFINING PARAMETERS

age_at_maturity <- 6 # SRC: Boyle - skeletochronology
adult_fem_abundance <- 1840 # SRC: Moldowan, estimated in 2019
s_a <- 1 / (1 + exp(-0.9656941)) # SRC: Dylan's intercept
f <- 1.41 # we know Patrick's estimates are underestimates, so works nicely with an overestimated p_skip
p_skip <- 0.5
s_m <- 0.8551498 # optimized from other parameters such that lambda = 1

get_rescaled_vector <- function(some_vector, k) {
  n_stable_fem <- some_vector[length(some_vector)]
  factor <- k / n_stable_fem
  return(factor * some_vector)
}

# CONSTRUCTING PROJECTION MATRIX

dimension <- age_at_maturity + 1 # ages 0-1, ..., 6-7, and 7+
model1_matrix <- matrix(0, dimension, dimension)
model1_matrix[1, dimension] <- f*(1-p_skip)
model1_matrix[dimension, dimension] <- s_a
for (i in 1:(dimension - 1)) {
  model1_matrix[i+1, i] <- s_m
}

eigen_info <- eigen.analysis(model1_matrix)
lambda1 <- eigen_info$lambda1
r1 <- log(lambda1)
stable_distribution <- eigen_info$stable.stage
stable_distribution_cal <- get_rescaled_vector(stable_distribution,
                                               adult_fem_abundance)

# SIMULATION HELPER FUNCTIONS

get_variant <- function(s) {
  result <- model1_matrix
  result[dimension, dimension] <- s
  return(result)
}


```

## 2. Simulation

```{r}

# setup
set.seed(416)
years_considered <- 2000:2100
envData <- final_df %>% filter(year %in% years_considered)

# SIMULATION FUNCTION

data_used <- envData
project_population <- function(scen) {
  # filter current scenario
  filteredData <- data_used %>% filter(scenario == scen) %>% arrange(year)
  # initialize pop vector
  init_N <- stable_distribution_cal
  # simulation setup
  curr_N <- init_N
  survivorships <- filteredData$survivorship
  pop_data <- numeric(length(years_considered))
  pop_data[1] <- curr_N[dimension]
  #simulation
  for (i in 1:(length(years_considered)-1)) {
    # update matrix with adult female survival
    curr_model_variant <- get_variant(survivorships[i])
    # do the simulatuion as normal
    curr_N <- curr_model_variant %*% curr_N
    # save the number of adult females in the current population vector
    pop_data[i+1] <- curr_N[dimension]
  }
  return(pop_data)
}

# PLOTTING

plotting_data <- data.frame(x_axis = years_considered,
                            y1 = project_population("SSP585"))


plotting_data %>% ggplot(aes(x = x_axis)) +
  geom_line(aes(y = y1, colour = "RCP 8.5")) +
  scale_y_continuous(limits = c(0, NA)) +
  geom_vline(xintercept = 2020, linetype = "dashed", colour = "black") +
  theme_classic() +
  labs(x = "Time (Year)", y = "Adult Female Abundance", colour = "Scenario")

```
