```{r}
# libraries
library(tidyverse)
library(ncdf4)
library(lubridate)
```

**Website Information:**

-   URL: "<https://climate-scenarios.canada.ca/?page=CanDCS6-data>"

-   Dataset: CanDCS-M6 (chose over U6 because better preseves relationships between precipitation and temperature ; statistically downscaled from CMIP6)

-   Variable: Mean temp / Min temp / Total precip

-   Model: Ensemble

-   Ensemble: Ensemble Mean

-   Scenario: Historical + SSP1-26 / SSP2-45 / SSP5-85

-   East Gate Climate Station Decimal Coordinates (NE): (`45.5475`, `-78.2464`)

-   East Gate Climate Station Decimal Coordinates (SW): (`45.5191`, `-78.2869`)

-   Time of Year: March / April / May

-   Start Year: 2000

-   End Year: 2100

-   Temporal Resolution: Daily output

**Protocol**

This resuts in 27 individual NC files and 9 unique scenario-variable combinations. Below, we:

1)  transform each of these into a dataframe

2)  bind all the dataframes together into single file, to be used for subsequent analysis

## NC Parsing Function

```{r}

# This code chunk contains the parsing function (and associated helpers) required to convert .nc files into R-compatable dataframes. Notes:
  # <ncFile> must only have ONE spatial grid point
  # <varname> must be a string
  # <scenario> must be an integer

convert_nc_to_df <- function(ncConnection, scenarioName, varName) {
  # extract nc data, then close connection
  time     <- ncvar_get(ncConnection, "time")
  currVar  <- ncvar_get(ncConnection, varName)
  lat      <- ncvar_get(ncConnection, "lat")
  lon      <- ncvar_get(ncConnection, "lon")
  if (length(c(lat, lon)) != 2) {
    error("Somehow, more than one grid point was accessed.")
  }
  nc_close(ncConnection)
  dates <- as.Date(time, origin = "1850-01-01")
  # put into dataframe, then return
  df <- data.frame(
    date = dates,
    year = year(dates),
    month = month(dates, label = TRUE),
    day = day(dates),
    ordinalDay = yday(dates),
    varName = varName,
    varValue = currVar,
    scenario = scenarioName
    )
  message(paste("Successfully parsed a NetCDF connection"))
  return(df)
}
```

## NC Loading Function

```{r}
# This function loads all of the NC files in a folder, and places them in a list object in R. Notes:
  # requires that all files have the same scenario AND variable

load_all_nc_files <- function(folder_path) {
  # get list of all nc files (full system paths)
  file_list <- list.files(path = folder_path, 
                          pattern = "\\.nc$", 
                          full.names = TRUE)
  # open them all, and name them based on their filenames
  nc_objects <- lapply(file_list, nc_open)
  names(nc_objects) <- basename(file_list)
  # return
  message(paste("Successfully loaded", length(nc_objects), "NetCDF connections."))
  return(nc_objects)
}
```

## Execution

```{r}

# setup
generic_filepath <- "/Users/benedictcummins-mburu/Library/CloudStorage/OneDrive-UniversityofToronto/Desktop/UofT/EEB398/Salamander Model/Official Code & Datasets/NC_"
scenario_list <- c("SSP585", "SSP245", "SSP126")
var_list <- c("tas", "tasmin", "pr")
ECCclimateData <- data.frame(
    date = c(), # date object, represented as a string of the form: YYYY-MM-DD
    year =  integer(),
    month = character(),
    day = integer(),
    ordinalDay = numeric(),
    varName = character(),
    varValue = numeric(),
    scenario = character()
    )
# populate df with NC data from all files 
for (scenario in scenario_list) {
  for (var in var_list) {
    curr_path <- paste0(generic_filepath, var, scenario)
    curr_allNC <- load_all_nc_files(curr_path)
    currDataList <- lapply(curr_allNC , convert_nc_to_df,
                           scenarioName = scenario, varName = var)
    for (df in currDataList) {
      ECCclimateData <- rbind(ECCclimateData, df)
    }
  }
}
```

```{r}
# pivot wider to condense rows
savedData <- pivot_wider(ECCclimateData, names_from = varName, values_from = varValue)
# uncomment to write to CSV
# write.csv(savedData, "EC_ClimateProjections.csv", row.names = FALSE)
```

## Post-Hoc Testing {style="color:blue"}

1.  Before 2023, trajectories for different scenarios were similar, but not exactly the same.

```{r}

savedData <- read.csv("EC_ClimateProjections.csv")
yearCheck <- 2021

month_order <- c("Mar", "Apr", "May")

# 2. Convert and Arrange
savedData <- savedData %>%
  mutate(month = factor(month, levels = month_order)) %>%
  arrange(year, month)

thing1 <- savedData %>% filter(year == yearCheck) %>% filter(scenario == "SSP126")
thing2 <- savedData %>% filter(year == yearCheck) %>% filter(scenario == "SSP585")
thing3 <- savedData %>% filter(year == yearCheck) %>% filter(scenario == "SSP245")
plot(x = thing1$ordinalDay, y = thing1$tasmin, col = "forestgreen", type = "l")
lines(x = thing2$ordinalDay, y = thing2$tasmin, col = "darkred")
lines(x = thing3$ordinalDay, y = thing3$tasmin, col = "yellow")
```

Outputted gridpoint: 45.54167, -78.29166
