# Libraries + Helper Functions

```{r}
# libraries 
library(tidyverse)
library(viridis)
library(ncdf4)
library(lubridate)
library(readxl)
library(popbio)

# rescale a stable-stage vector for a given adult female carrying capacity k
get_rescaled_vector <- function(some_vector, k) {
  n_stable_fem <- some_vector[length(some_vector) - 1]
  factor <- k / n_stable_fem
  return(factor * some_vector)
}

# inverse logit transform
inverse_logit_transform <- function(x) {
  return(1 / (1 + exp(-x)))
}
```

# Data Cleaning

## Cleaning EnvCan Raw Data

**Website Information:**

-   URL: "<https://climate-scenarios.canada.ca/?page=CanDCS6-data>"

-   Dataset: CanDCS-U6 (downscaling method)

-   Model: Ensemble

-   Ensemble: Ensemble Mean

-   Bat Lake Coordinates (NE): (45.58671, -78.5182)

-   Bat Lake Coordinates (SW): (45.58469, -78.5206)

**Climate Scenarios Considered:**

-   1 = SSP1-26

-   2 = SSP2-45

-   3 = SSP5-85

In `CLEANING - MEAN TEMPS`, we create a new CSV file ("ENVCAN_climate_projections_means.csv") containing the projected mean daily temperature every March, April, and May, from 2025 until 2100. This is done for the 3 climate scenarios considerered above.

```{r ENV CANADA DATA - EXTRACTING MEAN SPRING TEMPS, eval = FALSE}

# READ RAW DATA
nc_mar_1 <- nc_open("bccaqv2r_tas_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_1 <- nc_open("bccaqv2r_tas_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_1 <- nc_open("bccaqv2r_tas_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_2 <- nc_open("bccaqv2r_tas_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_2 <- nc_open("bccaqv2r_tas_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_2 <- nc_open("bccaqv2r_tas_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_3 <- nc_open("bccaqv2r_tas_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_3 <- nc_open("bccaqv2r_tas_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_3 <- nc_open("bccaqv2r_tas_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")

# DATA EXTRACTION
extract_raw_climate_data_mean <- function(nc_data) {
  # get necessary nc data, then close the file
  time <- ncvar_get(nc_data, "time")
  temp <- ncvar_get(nc_data, "tas")
  nc_close(nc_data)
  # convert dates
  dates <- as.Date(time, origin = "1850-01-01")
  # get scenario
  nc_name <- deparse(substitute(nc_data))
  scenario_index <- substr(nc_name, nchar(nc_name), nchar(nc_name))
  # build dataframe
  df <- data.frame(year = year(dates),
                   month = month(dates),
                   day = day(dates),
                   ordinal_day = yday(dates),
                   avg_temp = temp,
                   scenario = scenario_index)
  return(df)
}

df_mar_1 <- extract_raw_climate_data_mean(nc_mar_1)
df_apr_1 <- extract_raw_climate_data_mean(nc_apr_1)
df_may_1 <- extract_raw_climate_data_mean(nc_may_1)
df_mar_2 <- extract_raw_climate_data_mean(nc_mar_2)
df_apr_2 <- extract_raw_climate_data_mean(nc_apr_2)
df_may_2 <- extract_raw_climate_data_mean(nc_may_2)
df_mar_3 <- extract_raw_climate_data_mean(nc_mar_3)
df_apr_3 <- extract_raw_climate_data_mean(nc_apr_3)
df_may_3 <- extract_raw_climate_data_mean(nc_may_3)

final_df <- bind_rows(df_mar_1, df_apr_1, df_may_1,
                      df_mar_2, df_apr_2, df_may_2,
                      df_mar_3, df_apr_3, df_may_3)

# uncomment this line to write the CSV file to your local directory
write.csv(final_df, "ENVCAN_climate_projections_means.csv", row.names = FALSE)
```

In `CLEANING - MIN TEMPS`, we create a new CSV file ("ENVCAN_climate_projections_mins.csv") containing the projected minimum daily temperature every March, April, and May, from 2025 until 2100. This is done for the 3 climate scenarios considerered above.

```{r CLEANING - MIN TEMPS, eval = FALSE}

# READ IN RAW DATA
nc_mar_1 <- nc_open("bccaqv2r_tasmin_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_1 <- nc_open("bccaqv2r_tasmin_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_1 <- nc_open("bccaqv2r_tasmin_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_2 <- nc_open("bccaqv2r_tasmin_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_2 <- nc_open("bccaqv2r_tasmin_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_2 <- nc_open("bccaqv2r_tasmin_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_3 <- nc_open("bccaqv2r_tasmin_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_3 <- nc_open("bccaqv2r_tasmin_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_3 <- nc_open("bccaqv2r_tasmin_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")

# DATA EXTRACTION FUNCTION
extract_raw_climate_data_min <- function(nc_data) {
  # get necessary nc data
  time <- ncvar_get(nc_data, "time")
  temp <- ncvar_get(nc_data, "tasmin")
  nc_close(nc_data)
  # convert dates
  dates <- as.Date(time, origin = "1850-01-01")
  # get scenario
  nc_name <- deparse(substitute(nc_data))
  scenario_index <- substr(nc_name, nchar(nc_name), nchar(nc_name))
  # build dataframe
  df <- data.frame(year = year(dates),
                   month = month(dates),
                   day = day(dates),
                   ordinal_day = yday(dates),
                   min_temp = temp,
                   scenario = scenario_index)
  return(df)
}

df_mar_1 <- extract_raw_climate_data_min(nc_mar_1)
df_apr_1 <- extract_raw_climate_data_min(nc_apr_1)
df_may_1 <- extract_raw_climate_data_min(nc_may_1)
df_mar_2 <- extract_raw_climate_data_min(nc_mar_2)
df_apr_2 <- extract_raw_climate_data_min(nc_apr_2)
df_may_2 <- extract_raw_climate_data_min(nc_may_2)
df_mar_3 <- extract_raw_climate_data_min(nc_mar_3)
df_apr_3 <- extract_raw_climate_data_min(nc_apr_3)
df_may_3 <- extract_raw_climate_data_min(nc_may_3)

final_df <- bind_rows(df_mar_1, df_apr_1, df_may_1,
                      df_mar_2, df_apr_2, df_may_2,
                      df_mar_3, df_apr_3, df_may_3)

# uncomment this line to write the CSV file to your local directory
write.csv(final_df, "ENVCAN_climate_projections_mins.csv", row.names = FALSE)

```

In `CLEANING - TOTAL PRECIP`, we create a new CSV file ("ENVCAN_climate_projections_precip.csv") containing the projected total daily precipitation every March, April, and May, from 2025 until 2100. This is done for the 3 climate scenarios considerered above.

```{r CLEANING - TOTAL PRECIP, eval = FALSE}

# READ IN RAW DATA
nc_mar_1 <- nc_open("bccaqv2r_pr_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_1 <- nc_open("bccaqv2r_pr_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_1 <- nc_open("bccaqv2r_pr_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_2 <- nc_open("bccaqv2r_pr_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_2 <- nc_open("bccaqv2r_pr_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_2 <- nc_open("bccaqv2r_pr_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")
nc_mar_3 <- nc_open("bccaqv2r_pr_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_mar.nc")
nc_apr_3 <- nc_open("bccaqv2r_pr_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_apr.nc")
nc_may_3 <- nc_open("bccaqv2r_pr_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2025_2100_may.nc")

# DATA EXTRACTION FUNCTION
extract_raw_climate_data_precip <- function(nc_data) {
  # get necessary nc data
  time <- ncvar_get(nc_data, "time")
  precip <- ncvar_get(nc_data, "pr")
  nc_close(nc_data)
  # convert dates
  dates <- as.Date(time, origin = "1850-01-01")
  # get scenario number
  nc_name <- deparse(substitute(nc_data))
  scenario_index <- substr(nc_name, nchar(nc_name), nchar(nc_name))
  # build dataframe
  df <- data.frame(year = year(dates),
                   month = month(dates),
                   day = day(dates),
                   ordinal_day = yday(dates),
                   total_precip = precip,
                   scenario = scenario_index)
  return(df)
}

df_mar_1 <- extract_raw_climate_data_precip(nc_mar_1)
df_apr_1 <- extract_raw_climate_data_precip(nc_apr_1)
df_may_1 <- extract_raw_climate_data_precip(nc_may_1)
df_mar_2 <- extract_raw_climate_data_precip(nc_mar_2)
df_apr_2 <- extract_raw_climate_data_precip(nc_apr_2)
df_may_2 <- extract_raw_climate_data_precip(nc_may_2)
df_mar_3 <- extract_raw_climate_data_precip(nc_mar_3)
df_apr_3 <- extract_raw_climate_data_precip(nc_apr_3)
df_may_3 <- extract_raw_climate_data_precip(nc_may_3)

final_df <- bind_rows(df_mar_1, df_apr_1, df_may_1,
                      df_mar_2, df_apr_2, df_may_2,
                      df_mar_3, df_apr_3, df_may_3)

# uncomment this line to write the CSV file to your local directory
write.csv(final_df, "ENVCAN_climate_projections_precip.csv", row.names = FALSE)

```

## Final Data Cleaning

**Ultimate Goal**: produce a dataset where for each year-scenario pair from 2025 - 2100 we have estimated (i) average minimum temperature of the breeding season, and (ii) precipitation (measured by calculating the total precipitation over a 21-day interval centered around peak egg mass date. From these two estimates, we use the survivorship model to estimate the average adult female survivorship for that year-scenario pair.

**Some Information:**

-   The breeding season is defined to start on the Bat Lake (BL) ice out date, and end at peak egg mass date.

-   We used mean April temperature to predict Bat Lake ice out date using simple linear regression.

-   Then, we assumed that peak egg mass date always occurs 22 days after ice out (corresponding to the average observed breeding season length during the 2008-2022 BLISS study period). Using these two estimates, we can easily construct both the breeding interval as well as the precipitation interval.

-   Using these intervals along with daily ENVCAN estimates of minimum temperature and total precipitation, we calculated (i) and (ii) for each year-scenario pair.

-   *Note: all dates are measured and used ordinally*

**Regression information:**

-   BL ice out date is predicted well ($R^2 = 0.5$) by mean April temperature (Note: estimates taken from a screenshot of the summary output of a fitted simple regression model).

```{r CLEANING - OTHER DATASETS USED}

BLISS_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
```

```{r CLEANING - HELPER FUNCTIONS}

# read in cleaned ENVCAN data
projected_data_means <- read.csv("ENVCAN_climate_projections_means.csv")
projected_data_mins <- read.csv("ENVCAN_climate_projections_mins.csv")
projected_data_precip <- read.csv("ENVCAN_climate_projections_precip.csv")

precip_threshold <- 0.1 # SRC: minimum reprorted daily value in the East Gate Climate dataset

# using ENVCAN data, estimate the mean of daily average temperatures in April in "year" and for scenario = "scenario"
get_mean_april_temp <- function(this_year, this_scenario) {
  filtered_data <- projected_data_means %>% filter(year == this_year,
                                                   scenario == this_scenario,
                                                   month == 4)
  return(mean(filtered_data$avg_temp, na.rm = T))
}

# LEGACY
# using simple linear regression, estimate LOTR ice out date from mean April temperature
get_LOTR_ice_out_from_april_temp <- function(mean_april_temp) {
  # SOURCE: Dylan did a regression (iMessage screenshot)
  m <- -2.9912
  b <- 125.9639
  return((m*mean_april_temp) + b)
}

# LEGACY
# using simpe linear regression, estimate BL ice out date from LOTR ice out date
get_BL_ice_out_from_LOTR_ice_out <- function(ordinal_date) {
  # SOURCE: Dylan did a regression (e-mail screenshot)
  m <- 0.68665
  b <- 37.52535
  return((m*ordinal_date) + b)
}

# using simpe linear regression, estimate BL ice out date from mean April temperature
get_BL_ice_out_from_april_temp <- function(ordinal_date) {
  # SOURCE: Dylan did a regression (iMessage screenshot)
  m <- -3.1373
  b <- 125.5190
  return((m*ordinal_date) + b)
}


# using ENVCAN minimum daily temperature data + breeding interval estimates, estimate the mean minimum daily temperature during the breeding season for that year-scenario pair
get_avg_min_temp <- function(this_year, this_scenario, start, end) {
  # filter the data as required
  filtered_data <- projected_data_mins %>% filter(year == this_year,
                                                  scenario == this_scenario,
                                                  ordinal_day >= start,
                                                  ordinal_day <= end)
  # return average of min temps within interval
  return(mean(filtered_data$min_temp))
}

# using ENVCAN total daily precipitation data + precipitation interval estimates, estimate the precipitation measure for that year-scenario pair
get_precip <- function(this_year, this_scenario, start, end) {
  # filter and mutate the data as required
  filtered_data <- projected_data_precip %>% filter(year == this_year,
                                                    scenario == this_scenario,
                                                    ordinal_day >= start,
                                                    ordinal_day <= end) %>%
    mutate(precip = ifelse(total_precip >= precip_threshold, 1, 0),
           total_precip = ifelse(precip, total_precip, 0))
  t1 <- sum(filtered_data$total_precip) # get the total precipitation over entire interval
  t2 <- sum(filtered_data$precip) # get the total number of days with non-zero rain
  # return value
  if(t2 == 0) return(0)  # to avoid division by 0 error
  return((t1 / t2) / 21)
}

get_precip <- function(this_year, this_scenario, start, end) {
  # filter and mutate the data as required
  filtered_data <- projected_data_precip %>% filter(year == this_year,
                                                    scenario == this_scenario,
                                                    ordinal_day >= start,
                                                    ordinal_day <= end) %>%
    mutate(precip = ifelse(total_precip >= precip_threshold, 1, 0),
           total_precip = ifelse(precip, total_precip, 0))
  t1 <- sum(filtered_data$total_precip) # get the total precipitation over entire interval
  t2 <- sum(filtered_data$precip) # get the total number of days with non-zero rain
  # return value
  if(t2 == 0) return(0)  # to avoid division by 0 error
  return((t1 / t2) / 21)
}
```

```{r CLEANING - CREATE FINAL DATASET}

# read in cleaned ENVCAN data
projected_data_means <- read.csv("ENVCAN_climate_projections_means.csv")
projected_data_mins <- read.csv("ENVCAN_climate_projections_mins.csv")
projected_data_precip <- read.csv("ENVCAN_climate_projections_precip.csv")

# setup
n_years <- 2100 - 2025 + 1
mean_breeding_season_length <- round(mean(BLISS_study_data$Breeding.Season.Length, na.rm = T))
precipitation_interval_length <- 21

# initialize interval dataframe (Note: interval bounds measured in ordinal days)
intervals <- data.frame(year = integer(),
                        scenario = integer(),
                        breed_start = integer(),
                        breed_end = integer(),
                        precip_start = integer(),
                        precip_end = integer(),
                        mean_April_temp = numeric())

# estimating interval lengths for each year-scenario pair
for (scenario in 1:3) {
  # initialize empty vectors
  curr_breed_starts <- numeric(n_years)
  curr_breed_ends <- numeric(n_years)
  curr_precip_starts <- numeric(n_years)
  curr_precip_ends <- numeric(n_years)
  curr_projected_breeding_ends <- numeric(n_years)
  curr_mean_April_temps <- numeric(n_years)
  # for each year from 2025-2100...
  for (year in 2025:2100) {
  # get estimated mean April temperature from ENVCAN data
  curr_April_temp <- get_mean_april_temp(year, scenario)
  # estimate BL ice out date from mean April temp
  curr_BL_date <- round(get_BL_ice_out_from_april_temp(curr_April_temp))
  # assume peak egg mass date is 22 - 1 days later
  curr_peak_egg_mass_date <- curr_BL_date + mean_breeding_season_length - 1
  # save estimated interval lengths
  i <- year - 2025 + 1
  curr_breed_starts[i] <- curr_BL_date
  curr_breed_ends[i] <- curr_peak_egg_mass_date
  curr_precip_starts[i] <- curr_peak_egg_mass_date - 10
  curr_precip_ends[i] <- curr_peak_egg_mass_date + 10
  curr_mean_April_temps[i] <- curr_April_temp
  }
  # save results in a temporary dataframe to then be appended to the main
  temp_df <- data.frame(
    year = 2025:2100,
    scenario = scenario,
    breed_start = curr_breed_starts,
    breed_end = curr_breed_ends,
    precip_start = curr_precip_starts,
    precip_end = curr_precip_ends,
    mean_April_temp = curr_mean_April_temps)
  intervals <- rbind(intervals, temp_df)
}

# initialize final dataframe
final_data <- data.frame(year = integer(),
                         scenario = integer(),
                         breed_start = integer(),
                         breed_end = integer(),
                         precip_start = integer(),
                         precip_end = integer(),
                         avg_min_temp = numeric(),
                         precipitation = numeric(),
                         mean_April_temp = numeric())
# populate final dataframe
for (j in 1:nrow(intervals)) {
  curr_avg_min_temp <- get_avg_min_temp(intervals[j, 1], intervals[j, 2],
                                        intervals[j, 3], intervals[j, 4])
  curr_precip <- get_precip(intervals[j, 1], intervals[j, 2],
                            intervals[j, 5], intervals[j, 6])
  temp_df1 <- data.frame(year = intervals[j, 1],
                         scenario = intervals[j, 2],
                         breed_start = intervals[j, 3],
                         breed_end = intervals[j, 4],
                         precip_start = intervals[j, 5],
                         precip_end = intervals[j, 6],
                         avg_min_temp = curr_avg_min_temp,
                         precipitation = curr_precip,
                         mean_April_temp = intervals[j, 7])
  final_data <- rbind(final_data, temp_df1)
}

# uncomment to read final df to CSV file
write.csv(final_data, "final_data2.csv")
```

# Simulations

## Model Specification

Using female-only Leslie matrix (may extend to a logistic-type density-dependent model version), but where adult female survivorship updates at each timestep according to the projected climate variables of that year.

Note: simulation trajectories are robust to the choice of p_skip. P-skip only matters insofar as it determines how reasonable our metamorph survival estimate is (can write code that shows this to be true, but for now it's TMB).

```{r SIMULATION - BASE MODEL PARAMETERS}

age_at_maturity <- 6 # SRC: Boyle - skeletochronology
adult_fem_abundance <- 1840 # SRC: Moldowan, estimated in 2019
s_a <- inverse_logit_transform(0.9656941) # mean adult female survival (Dylan's intercept)
f <- 1.41 # we know Patrick's estimates are underestimates, so works nicely with an overestimated p_skip
p_skip <- 0.5
s_m <- 0.8551498 # optimized from other parameters such that lambda = 1

# other parameters
p_skip2 <- 1 - 0.32 # SRC: Husting
s_m2 <- 0.9072593
```

```{r SIMULATION - MODEL MATRIX + EIGEN STUFF}

dimension <- age_at_maturity + 1 # ages 0-1, ..., 6-7, and 7+
model1_matrix <- matrix(0, dimension, dimension)
model1_matrix[1, dimension] <- f*(1-p_skip)
model1_matrix[dimension, dimension] <- s_a
for (i in 1:(dimension - 1)) {
  model1_matrix[i+1, i] <- s_m
}

eigen_info <- eigen.analysis(model1_matrix)
lambda1 <- eigen_info$lambda1
r1 <- log(lambda1)
stable_distribution <- eigen_info$stable.stage
stable_distribution_cal <- get_rescaled_vector(stable_distribution,
                                                        adult_fem_abundance)
```

## Helper Functions

Adult female survivorship $\alpha$ is significantly related to (i) standardized average minimum breeding season tenmperature $X_1$, and (ii) a specific measure of total precipitation $X_2$ (total precip / \# days wtht rain / 21).

```{r SIMULATION - OTHER DATASETS USED}

unstandardized_BLISS_climate_data <- read.csv("Unstandardized_INB.T_OUT.P.csv")
unstandardized_BLISS_climate_data <- unstandardized_BLISS_climate_data[complete.cases(unstandardized_BLISS_climate_data), ]
```

```{r SIMULATION - HELPER FUNCTIONS}

# climate values observed during BLISS project
obs_temp_mean <- mean(unstandardized_BLISS_climate_data$T.INB)
obs_temp_sd <- sd(unstandardized_BLISS_climate_data$T.INB)
obs_precip_mean <- mean(unstandardized_BLISS_climate_data$P.OUT)
obs_precip_sd <- sd(unstandardized_BLISS_climate_data$P.OUT)

# using Dylan's model, get survival from mean minimum temperature AND precipitation measure
# Note: this function STANDARDIZES inputs (so no need to standardize inputs beforehand)
get_survival_from_key_metrics <- function(temp, precip) {
  # SOURCE: Dylan's model fitted values
  b0 <- 0.9656941 # intercept
  b1 <- -0.762042 # standardized temp
  b2 <- 0.4460179 # standardized precipitation
  # standardize temp and precip
  standardized_temp <- (temp - obs_temp_mean) / obs_temp_sd
  standardized_precip <- (precip - obs_precip_mean) / obs_precip_sd
  # estimate mean respionse using logit model
  logit_s <- b0 + b1 * standardized_temp + b2 * standardized_precip
  # convert back to raw probability measure
  s <- inverse_logit_transform(logit_s)
  # return result
  return (s)
}

# produce a model matrix where the base model's survivorship is adjusted according to specified temp and precip
get_variant <- function(temp, precip) {
  new_s_a <- get_survival_from_key_metrics(temp, precip)
  result <- model1_matrix
  result[dimension, dimension] <- new_s_a
  return(result)
}
```

## Projections

```{r POPULATION PROJECTIONS}

# setup
set.seed(416)
envData <- read.csv("final_data2.csv")
years_considered <- 2025:2100

# APPLY TEMPERATURE CORRECTION (LEGACY)
corrected_env_data <- envData %>% mutate(avg_min_temp = ifelse(
  scenario == 1, avg_min_temp - 2.282 + 0.953, ifelse(
    scenario == 2, avg_min_temp - 2.30 + 0.953, avg_min_temp - 2.279 + 0.953)))

# MEAN-SHIFTED DATA
meanShift_avgMinTemp <- -1.281596
meanShift_precipitation <- 0.1453267
meanShift_envData <- envData %>% mutate(
  avg_min_temp = avg_min_temp + meanShift_avgMinTemp,
  precipitation = precipitation + meanShift_precipitation)

# FULL-SCALE DATA

# SIMULATION FUNCTION
data_used <- meanShift_envData
project_population <- function(scenario_number) {
  # filter env_data to current scenario
  curr_env_data <- data_used %>%
    filter(scenario == scenario_number)
  # initialize pop vector at stable distribution (rescaled to estimated adult female abundance)
  init_N <- stable_distribution_cal
  # simulation setup
  curr_N <- init_N
  survivorships <- numeric(length(years_considered))
  survivorships[1] <- model1_matrix[dimension, dimension]
  pop_data <- numeric(length(years_considered))
  pop_data[1] <- curr_N[dimension]
  #simulation
  for (i in 1:(length(years_considered)-1)) {
    # update adult female survival according to 2 environmental metrics
    curr_mean_min_temp <- curr_env_data$avg_min_temp[i+1]
    curr_precip <- curr_env_data$precipitation[i+1]
    curr_model_variant <- get_variant(curr_mean_min_temp, curr_precip)
    # do the simulatuion as normal
    curr_N <- curr_model_variant %*% curr_N
    # save the number of adult females in the current population vector
    pop_data[i+1] <- curr_N[dimension]
    survivorships[i+1] <- curr_model_variant[dimension, dimension]
  }
  # return result
  return(pop_data)
}

# PLOTTING

plotting_data <- data.frame(x_axis = years_considered,
                            y1 = project_population(1),
                            y2 = project_population(2),
                            y3 = project_population(3) )


plotting_data %>% ggplot(aes(x = x_axis)) +
  geom_line(aes(y = y1, colour = "RCP 2.6")) +
  geom_line(aes(y = y2, colour = "RCP 4.5")) +
  geom_line(aes(y = y3, colour = "RCP 8.5")) +
  scale_y_continuous(limits = c(0, NA)) +
  theme_classic() +
  labs(x = "Time (Year)", y = "Adult Female Abundance", colour = "Scenario")
```

# Exploratory Data Analysis (EXTRA)

## Comparing Observed vs. Predicted Climate Data

-   Regardless of the scenario, we observe continuity in the ENVCAN historic and predicted data.

-   

```{r GET PREDICTED MINIMUM SPRING TEMPS 2008-2022, eval = FALSE}

# PARAMETERS TO PUT INTO WEBSITE
url <- "https://climate-scenarios.canada.ca/?page=CanDCS6-data"
NE_bat_lake_coords <- c(45.58671, -78.5182)
SW_bat_lake_coords <- c(45.58469, -78.5206)
# 1 = SSP1-26, 2 = SSP2-45, 3 = SSP5-85
# gets the min temperature from 2008 - 2022 in Apr and May, all scenarios

# WEBSITE RAW OUTPUT
nc_apr_1 <- nc_open("bccaqv2r_tasmin_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_apr.nc")
nc_may_1 <- nc_open("bccaqv2r_tasmin_ssp126_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_may.nc")
nc_apr_2 <- nc_open("bccaqv2r_tasmin_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_apr.nc")
nc_may_2 <- nc_open("bccaqv2r_tasmin_ssp245_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_may.nc")
nc_apr_3 <- nc_open("bccaqv2r_tasmin_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_apr.nc")
nc_may_3 <- nc_open("bccaqv2r_tasmin_ssp585_ensemblemean_78.518W_45.670N_78.604W_45.585N_2008_2022_may.nc")

# DATA EXTRACTION
extract_raw_climate_data_hist_min <- function(nc_data) {
  # get necessary nc data
  time <- ncvar_get(nc_data, "time")
  temp <- ncvar_get(nc_data, "tasmin")
  nc_close(nc_data)
  # convert dates
  dates <- as.Date(time, origin = "1850-01-01")
  # get scenario
  nc_name <- deparse(substitute(nc_data))
  scenario_index <- substr(nc_name, nchar(nc_name), nchar(nc_name))
  # build dataframe
  df <- data.frame(year = year(dates),
                   month = month(dates),
                   day = day(dates),
                   min_temp = temp,
                   scenario = scenario_index)
  return(df)
}

df_apr_1 <- extract_raw_climate_data_hist_min(nc_apr_1)
df_may_1 <- extract_raw_climate_data_hist_min(nc_may_1)
df_apr_2 <- extract_raw_climate_data_hist_min(nc_apr_2)
df_may_2 <- extract_raw_climate_data_hist_min(nc_may_2)
df_apr_3 <- extract_raw_climate_data_hist_min(nc_apr_3)
df_may_3 <- extract_raw_climate_data_hist_min(nc_may_3)

final_df <- bind_rows(
  df_apr_1, df_may_1,
  df_apr_2, df_may_2,
  df_apr_3, df_may_3
)
# write.csv(final_df, "ENVCAN_historic_mins.csv", row.names = FALSE)
```

```{r COMPARE HISTORIC VS. PREDICTED MIN TEMP, warning = FALSE, eval = FALSE}
library(zoo)

# load in data
historic_mins <- read.csv("ENVCAN_historic_mins.csv")
predicted_mins <- read.csv("ENVCAN_climate_projections_mins.csv") %>% filter(scenario == 3,
                                                                             month != 3)
full <- rbind(historic_mins, predicted_mins %>% select(-ordinal_day)) %>%
  mutate(date = as.Date(paste(year, month, day, sep = "-"))) %>% arrange(date) %>%
  mutate(period = ifelse(year <= 2022, "Historic", "Projected"),
         roll365 = rollmean(min_temp, k = 365, fill = NA, align = "center"))

# daily min temp over time
ggplot(full, aes(x = date, y = min_temp)) + geom_line(alpha = 0.4) + geom_line(aes(y = roll365, colour = period), linewidth = 1)
```

## Comparing Final Dataset to Observed Data

```{r}
# read in observed data
BL_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
BL_climate_data <- read.csv("Unstandardized_INB.T_OUT.P.csv")
BL_climate_data <- BL_climate_data[complete.cases(BL_climate_data), ]

# read in predicted data
scenario_chosen <- 3
predicted_data <- read.csv("final_data2.csv") %>% filter(scenario == scenario_chosen)

# setup
years <- 2008:2100

# mean April temperature
obs  <- rep(NA, 15)
pred <- predicted_data$mean_April_temp
mean_April_temp <- c(obs, NA, NA, pred)
cut <- length(obs) + 2
plot(years, mean_April_temp, type = "n")
lines(years[1:cut], mean_April_temp[1:cut], col = "black", lwd = 2)
lines(years[-(1:cut)], mean_April_temp[-(1:cut)], col = "red", lwd = 2)

# breed start
obs  <- BL_study_data$Ice.Out.Ordinal
pred <- predicted_data$breed_start
breed_start <- c(obs, NA, NA, pred)
plot(years, breed_start, type = "n")
lines(years[1:cut],  breed_start[1:cut],  col = "black", lwd = 2)
lines(years[-(1:cut)], breed_start[-(1:cut)], col = "red", lwd = 2)

# breed end
obs <- BL_study_data$Peak.Egg.Mass.Ordinal
pred <- predicted_data$breed_end
breed_end <- c(obs, NA, NA, pred)
plot(years, breed_end, type = "n")
lines(years[1:cut],  breed_end[1:cut],  col = "black", lwd = 2)
lines(years[-(1:cut)], breed_end[-(1:cut)], col = "red", lwd = 2)

# spring avg min temp
obs <- BL_climate_data$T.INB
pred <- predicted_data$avg_min_temp
avg_min_temp <- c(obs, NA, NA, pred)
plot(years, avg_min_temp, type = "n")
lines(years[1:cut],  avg_min_temp[1:cut],  col = "black", lwd = 2)
lines(years[-(1:cut)], avg_min_temp[-(1:cut)], col = "red", lwd = 2)

# precip
obs <- BL_climate_data$P.OUT
pred <- predicted_data$precipitation
precip <- c(obs, NA, NA, pred)
plot(years, precip, type = "n")
lines(years[1:cut],  precip[1:cut],  col = "black", lwd = 2)
lines(years[-(1:cut)], precip[-(1:cut)], col = "red", lwd = 2)

# survival
obs <- Map(get_survival_from_key_metrics, BL_climate_data$T.INB, BL_climate_data$P.OUT)
pred <- Map(get_survival_from_key_metrics, predicted_data$avg_min_temp, predicted_data$precipitation)
adult_fem_survival <- c(obs, NA, NA, pred)
plot(years, adult_fem_survival, type = "n")
lines(years[1:cut],  adult_fem_survival[1:cut],  col = "black", lwd = 2)
lines(years[-(1:cut)], adult_fem_survival[-(1:cut)], col = "red", lwd = 2)

```

Checked the ACF of all observed data, and it was all white noise (not even close). ACF of predicted mean April temp was not compatable with white noise. This means that we do not need to worry about interannual correlations in observed climate data.

## Applying Mean Shifting to Final Dataset

```{r}

# READ DATA 

# read in observed data
BL_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
BL_climate_data <- read.csv("Unstandardized_INB.T_OUT.P.csv")
BL_climate_data <- BL_climate_data[complete.cases(BL_climate_data), ]
# read in predicted data
scenario_chosen <- 3
predicted_data <- read.csv("final_data2.csv") %>% filter(scenario == scenario_chosen)

# MEAN-SHIFT CLIMATE PREDICTIONS
meanShift_avgMinTemp <- mean(BL_climate_data$T.INB) - predicted_data$avg_min_temp[1]
meanShift_precipitation <- mean(BL_climate_data$P.OUT) - predicted_data$precipitation[1]


# PLOTTING

# spring avg min temp
obs <- BL_climate_data$T.INB
pred1 <- predicted_data$avg_min_temp + meanShift_avgMinTemp
avg_min_temp <- c(obs, NA, NA, pred1)
plot(years, avg_min_temp, type = "n")
points(years, avg_min_temp, col = ifelse(is.na(avg_min_temp), NA,
       ifelse(1:length(avg_min_temp) <= length(obs) + 2, "black", "blue")))

# precip
obs <- BL_climate_data$P.OUT
pred2 <- predicted_data$precipitation + meanShift_precipitation
precip <- c(obs, NA, NA, pred2)
plot(years, precip, type = "n")
points(years, precip, col = ifelse(is.na(precip), NA,
       ifelse(1:length(precip) <= length(obs) + 2, "black", "blue")))

# survival
obs <- Map(get_survival_from_key_metrics, BL_climate_data$T.INB, BL_climate_data$P.OUT)
pred <- Map(get_survival_from_key_metrics, pred1, pred2)
adult_fem_survival <- c(obs, NA, NA, pred)
plot(years, adult_fem_survival, type = "n")
points(years, adult_fem_survival, col = ifelse(is.na(adult_fem_survival), NA,
       ifelse(1:length(adult_fem_survival) <= length(obs) + 2, "black", "blue")))
```

## Apply Full Scale to Final Dataset

```{r}

# READ DATA

# read in observed data
BL_study_data <- read_excel("BLISS_2008_2022_study_variables.xlsx")
BL_climate_data <- read.csv("Unstandardized_INB.T_OUT.P.csv")
BL_climate_data <- BL_climate_data[complete.cases(BL_climate_data), ]
# read in predicted data
scenario_chosen <- 2
predicted_data <- read.csv("final_data2.csv") %>% filter(scenario == scenario_chosen)

# MEAN-SHIFTED CLIMATE PREDICTIONS
meanShift_avgMinTemp <- mean(BL_climate_data$T.INB) - predicted_data$avg_min_temp[1]
meanShift_precipitation <- mean(BL_climate_data$P.OUT) - predicted_data$precipitation[1]
obs1_sd <- sd(BL_climate_data$T.INB)
obs2_sd <- sd(BL_climate_data$P.OUT)

# VARIANCE-INFLATING ALGORITHM
full_scale <- function(obsSD, meanShifted_pred) {
  # 1. Fit polynomial trend (degree = 2) to predicted data
  time <- 1:length(meanShifted_pred)
  time2 <- time**2
  trend_fit <- lm(meanShifted_pred ~ time + time2) # fit quadratic model
  trend_est <- predict(trend_fit)
  # 2. Compute standardized residuals of de-trended data
  residuals <- rstandard(trend_fit)
  # 3. Rescale predicted residuals to match observed SD
  residuals_scaled <- (residuals) * obsSD * 0.1
  # 4. Add residuals back to predicted trend
  pred_adjusted <- trend_est + residuals_scaled
  return(pred_adjusted)
}
# Assumption1: observed climate data is white noise (which is observed)
# Assumption2: trend can always be estimated well with quadratic regression

# PLOTTING

# spring avg min temp
obs1 <- BL_climate_data$T.INB
pred1 <- full_scale(obs1_sd, predicted_data$avg_min_temp + meanShift_avgMinTemp)
avg_min_temp <- c(obs, NA, NA, pred1)
plot(years, avg_min_temp, type = "n")
points(years, avg_min_temp, col = ifelse(is.na(avg_min_temp), NA,
       ifelse(1:length(avg_min_temp) <= length(obs) + 2, "black", "blue")))

# precip
obs2 <- BL_climate_data$P.OUT
pred2 <- full_scale(obs2_sd, predicted_data$precipitation + meanShift_precipitation)
precip <- c(obs, NA, NA, pred2)
plot(years, precip, type = "n")
points(years, precip, col = ifelse(is.na(precip), NA,
       ifelse(1:length(precip) <= length(obs) + 2, "black", "blue")))

# survival
obs <- Map(get_survival_from_key_metrics, BL_climate_data$T.INB, BL_climate_data$P.OUT)
pred <- Map(get_survival_from_key_metrics, pred1, pred2)
adult_fem_survival <- c(obs, NA, NA, pred)
plot(years, adult_fem_survival, type = "n")
points(years, adult_fem_survival, col = ifelse(is.na(adult_fem_survival), NA,
       ifelse(1:length(adult_fem_survival) <= length(obs) + 2, "black", "blue")))
```

# Other

## Optimizing Metamorph Survivorship

```{r OPTIMIZING METAMORPH SURVIVORSHIP, eval = FALSE}

# choose a matrix to optimize
which_matrix_to_optimize <- model1_matrix

# HELPER 1
create_alternate_matrix <- function(candidate_s_m) {
  alternate_matrix <- which_matrix_to_optimize
  for (i in 1:(dimension - 1)) {
  alternate_matrix[i+1, i] <- candidate_s_m
  }
  return(alternate_matrix)
}

# target function for optimization: returns distance from target
target_function <- function(candidate_s_m) {
  test_matrix <- create_alternate_matrix(candidate_s_m)
  test_lambda <- popbio::eigen.analysis(test_matrix)$lambda1
  return(abs(test_lambda - 1))
}

# optimization code
s_m_range <- c(0,1)
optimized_s_m <- optimize(target_function, interval = s_m_range)$minimum
optimized_s_m
metamorph_to_adult <- optimized_s_m ** age_at_maturity
metamorph_to_adult
```

## Plotting Models for Verification

```{r PLOT MODELS THEORETICALLY, eval = FALSE}

model_used <- model1_matrix

# SIMULATION
n_sims <- 100
pop_data <- numeric(n_sims)
data_used <- corrected_env_data
init_pop_vector <- get_calibrated_vector(abs(get_dom_eigenvector(model_used)), adult_fem_abundance, dimension)
curr_r <- log(abs(get_dom_eigenvalue(model_used)))
curr_K <- sum(init_pop_vector)
curr_pop_vector <- init_pop_vector

for (i in 1:n_sims) {
    # update adult female survival
    curr_model_variant <- get_variant(model_used, 1.5)
    # do the simulatuion as normal
    curr_pop_size <- sum(curr_pop_vector)
    curr_d <- exp((-curr_r*curr_pop_size)/curr_K)
    curr_pop_vector <- curr_d * (curr_model_variant %*% curr_pop_vector)
    # save the number of adult females in the current population vector
    pop_data[i] <- curr_pop_vector[dimension]
}

# PLOT
plotting_data <- data.frame(x_axis = 1:n_sims, y = pop_data)

plotting_data %>% ggplot(aes(x = x_axis, y = y)) +
  geom_line()
```

```{r PLOTTING FOR VERIFICATION, eval = FALSE}
# PLOT SPRING TEMP-SURVIVAL FUNCTION
spring_temps <- seq(-5, 15, by = 0.1)
survivals <- get_survival_from_spring_temp(spring_temps)
df <- data.frame(SpringTemp = spring_temps, Survival = survivals)
ggplot(df, aes(x = SpringTemp, y = Survival)) +
  geom_line() +
  labs(
    x = "Minimum Spring Temperature (°C)",
    y = "Predicted Survival Probability"
  )
```
